\paperTitle

### 2014-05-13 doc/icml2014/directed.tex
Minor fix to epsilon todo and uniform distribution todo.
Fixed epsilon related bug
Included new assumptions for conditional independence

% DONE: don't talk about parameter recovery yet, just establish exclusive views
%Finally, the claim we make is that we can recover parameters for any
  %graphical model where every hidden variable is a bottleneck:
Now define the following property:
\begin{property}[Uniformly bottlenecked]
  \label{prop:bottleneck}
  A clique $\sC$ is \emph{uniformly bottlenecked} if
  every hidden variable in $\sC$ is a bottleneck. 
  If every clique $\sC \in \sG$ is uniformly bottlenecked, then we say
  that the graph $\sG$ is uniformly bottlenecked as well.
\end{property}
Note that two hidden variables are allowed to have overlapping views.
%If every hidden variable is a bottleneck, then we can estimate $\mOpp{v}{i}$
It is clear that having uniform bottlenecks implies part (ii)
of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
but we will show that part (i) follows as well.

@@

Now we establish the link between bottlenecks and exclusive views
(see \appendixref{exclusive-views} for the complete proof):
\begin{lemma}[Uniformly bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
% PL: simplify to cliques (this is also bad notation, since \sG is a set of cliques, not nodes)
%Let $\sG' = \{ h_{i_1}, \dots, h_{i_m} \} \subseteq \sG$ be
  %a connected subgraph of hidden variables.
  %If every hidden variable
  %$h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive views property.
If a graphical model $\sG$ is uniformly bottlenecked,
then $\sG$ has the exclusive views property.
\end{lemma}
% PL: save space, doesn't add much
%\begin{proof}
%  The independence conditions required by the bottleneck condition
%  effectively guarantee that at least one of the three views for
%  a bottleneck be exclusive. 
%  \appendixref{exclusive-views} provides a complete proof.
%\end{proof}

@@

In this section, we provide formal conditions under which $\LearnMarginals$
will produce consistent estimates.
We let
$\mOpphat{v}{i}$,
$\hat Z_\sC$,
and $\hat M_\sV$,
denote estimators for
$\mOpp{v}{i}$,
$Z_\sC$,
and $M_\sV$,
respectively.

@@

%\todo{did they prove this; if so just say "they show", not "we can show"}
%Using results from
%  \citet{anandkumar12moments,anandkumar13tensor} we can show that
%  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
%  x_{v_2}, x_{v_3}$ has the following sample complexity,
%  \todo{define hat O}
%  \todo{max subscript is misaligned}
%\begin{align*}
%  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O\left( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} \right). 
%\end{align*}

From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  \assumptionref{non-degeneracy}, we have that $\mOpphat{v}{i}$
  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on the $k$-th singular value of
  $\mOpp{v}{i}$:
\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
  \label{thm:sample-complexity}
  Without loss of generality, let $v=1, i=1$, and let $x_1, x_2, x_3$ be three views for
  $h_1$. With probability at least $1 - \delta$,
\begin{align*}
  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
    &\le  \\
    &
  \hspace{-0.3in}
      O\left(
      % PL: remove pi_max because that's upper bounded by 1.
      \frac{k \log(1/\delta) } 
      {\sqrt{n} \, \pi\oft{1}_{\min} (\sigma_{k}(\mOpp{1}{1}) \sigma_{k}(\mOpp{1}{2}))^{5/2}} \right) \epsilon,
\end{align*}
where $\pi\oft{1}_{\min}$ is the smallest element of $\pi\oft{1}$.
\todo{what's $\epsilon$?}

%If $\sigma_{\min}(\Pr(c \given \Pa(h_c))) \le \sigma$ for every clique
%  $\sC \in \sG$ and hidden variable $h_c \in \sC$,
%\begin{align*}
%  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
%    &\le 
%      O\left( 
%      \frac{k \log(1/\delta) {\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
%      {\sqrt{n} \sigma^{5t}} \right) \epsilon,
%\end{align*}
%where $t$ is the length of the shortest topological ordering from $h_1$
%to the unique parent of $x_1$, $h_t$.
\end{theorem}

% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
Note that $\sigma_{k}(\mOpp{v}{i})$ can become extremely
small if $h_i$ and $x_v$ are connected via many intermediate hidden variables.
\todo{cite work - Sontag?}
%In general, $\sigma_{k}(\mOpp{v}{i})$ will depend exponentially on the
  %path length between the hidden variable $h_i$ and $x_v$.
%This is expected because the influence from $h_i$ to $x_v$
  %gets diluted for every latent variable it must factor through.

>>>>

% DONE: don't talk about parameter recovery yet, just establish exclusive views
%Finally, the claim we make is that we can recover parameters for any
  %graphical model where every hidden variable is a bottleneck:
Now define the following property:
\begin{property}[Uniformly bottlenecked]
  \label{prop:bottleneck}
  A clique $\sC$ is \emph{uniformly bottlenecked} if
  every hidden variable in $\sC$ is a bottleneck. 
  If every clique $\sC \in \sG$ is uniformly bottlenecked, then we say
  that the graph $\sG$ is uniformly bottlenecked as well.
\end{property}
Note that two hidden variables are allowed to have overlapping views.
%If every hidden variable is a bottleneck, then we can estimate $\mOpp{v}{i}$
It is clear that having uniform bottlenecks implies part (ii)
of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
but we will show that part (i) follows as well, given basic assumptions
on the conditional independence structure of the graphical model.

\begin{assumption}[Conditional Independence]
  \label{asm:ci}  
  \
  \begin{enumerate}
    \item If, conditioned on a variable $h$, two variables $x_1, x_2$
      are independent, then $x_1$ and $x_2$ are not both dependent on
      any other variable $h'$: $x_1 \perp x_2 \given h \implies x_1
      \perp h' \given h \vee x_2 \perp h' \given h$ for every $h' \in
      \sC$.
    \item
      For every clique $\sC \in \sG$, conditioned on a variable $h \in
      \sC$, all other variables in the clique are dependent:
      $h_i \not\perp h_j \given h$ for every $h_i, h_j \in \sC$.
    \item
      Given an active path in the unconditioned model, then it remains
        active when conditioned on $h$. 
      Let $a, b, c$ be three variables in the graph such that
      $$ a \not\perp b \wedge b \not\perp c \implies a \not\perp c,$$
      then,
      $$ a \not\perp b \given h \wedge b \not\perp c \given h \implies a \not\perp c \given h.$$
%      Suppose $$x \not\perp h_1 \wedge h_1 \not\perp h_2 \implies
%      x \not\perp h_2,$$ then $$x \not\perp h_1 \given h \wedge h_1 \not\perp h_2
%      \given h \implies x \not\perp h_2 \given h$$.
  \end{enumerate}
\end{assumption}

%TODO: Describe with example the cliques ruled out.

@@

Now we establish the link between bottlenecks and exclusive views
(see \appendixref{exclusive-views} for the complete proof):
\begin{lemma}[Uniformly bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
% PL: simplify to cliques (this is also bad notation, since \sG is a set of cliques, not nodes)
%Let $\sG' = \{ h_{i_1}, \dots, h_{i_m} \} \subseteq \sG$ be
  %a connected subgraph of hidden variables.
  %If every hidden variable
  %$h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive views property.
  If a graphical model $\sG$ satisfies conditional independence
  assumptions (\assumptionref{ci}) and is uniformly bottlenecked, then
  $\sG$ has the exclusive views property.
\end{lemma}
% PL: save space, doesn't add much
%\begin{proof}
%  The independence conditions required by the bottleneck condition
%  effectively guarantee that at least one of the three views for
%  a bottleneck be exclusive. 
%  \appendixref{exclusive-views} provides a complete proof.
%\end{proof}

@@

%In this section, we provide formal conditions under which $\LearnMarginals$
%will produce consistent estimates.
%We let
%$\mOpphat{v}{i}$,
%$\hat Z_\sC$,
%and $\hat M_\sV$,
%denote estimators for
%$\mOpp{v}{i}$,
%$Z_\sC$,
%and $M_\sV$,
%respectively.

@@

%\todo{did they prove this; if so just say "they show", not "we can show"}
%Using results from
%  \citet{anandkumar12moments,anandkumar13tensor} we can show that
%  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
%  x_{v_2}, x_{v_3}$ has the following sample complexity,
%  \todo{define hat O}
%  \todo{max subscript is misaligned}
%\begin{align*}
%  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O\left( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} \right). 
%\end{align*}
It is easy to see that $\LearnMarginals$ has a polynomial sample complexity because it composes two parts that individually have polynomial sample complexity.
From \citet{anandkumar13tensor} given the non-degeneracy assumptions
  \assumptionref{non-degeneracy}, we have that $\mOpphat{v}{i}$
  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
  that depends polynomially on the $k$-th singular value of
  $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
Note that $\sigma_{k}(\mOpp{v}{i})$ can become extremely
small if $h_i$ and $x_v$ are connected via many intermediate hidden variables.
\todo{cite work - Sontag?}
%\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
%  \label{thm:sample-complexity}
%  Without loss of generality, let $v=1, i=1$, and let $x_1, x_2, x_3$ be three views for
%  $h_1$. With probability at least $1 - \delta$,
%\begin{align*}
%  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
%    &\le  \\
%    &
%  \hspace{-0.3in}
%      O\left(
%      % PL: remove pi_max because that's upper bounded by 1.
%      \frac{k \log(1/\delta) } 
%      {\sqrt{n} \, \pi\oft{1}_{\min} (\sigma_{k}(\mOpp{1}{1}) \sigma_{k}(\mOpp{1}{2}))^{5/2}} \right),
%\end{align*}
%where $\pi\oft{1}_{\min}$ is the smallest element of $\pi\oft{1}$.

%If $\sigma_{\min}(\Pr(c \given \Pa(h_c))) \le \sigma$ for every clique
%  $\sC \in \sG$ and hidden variable $h_c \in \sC$,
%\begin{align*}
%  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
%    &\le 
%      O\left( 
%      \frac{k \log(1/\delta) {\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
%      {\sqrt{n} \sigma^{5t}} \right) \epsilon,
%\end{align*}
%where $t$ is the length of the shortest topological ordering from $h_1$
%to the unique parent of $x_1$, $h_t$.
%\end{theorem}
%In general, $\sigma_{k}(\mOpp{v}{i})$ will depend exponentially on the
  %path length between the hidden variable $h_i$ and $x_v$.
%This is expected because the influence from $h_i$ to $x_v$
  %gets diluted for every latent variable it must factor through.

### 2014-05-13 doc/icml2014/exclusive-views.tex
Included new assumptions for conditional independence

In \sectionref{directed}, we stated that bottlenecks guaranteed
  exclusive views. We provide the proof below.
  We generalize the lemma so that it applies
  on any connected subgraph, not just cliques.
  Also, the statement is made in terms of subgraphs,
  not the entire graph.

\begin{lemma*}[Bottlenecks guarantee exclusive views]
  %\label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \ldots, h_{i_m} \} \subseteq \sG$ be
  a connected subgraph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive view property.
\end{lemma*}

\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not.
  Note that this proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sG'$, let $\sV_{h}$ be the set of
  views for it and let ${\sG'}^- \eqdef \sG' \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'}$ for all $h'
  \in {\sG'}^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in {\sG'}^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
  exists some $h_1, h_2 \in {\sG'}^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected to $h_2$.
  By participation in the subgraph, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

  Thus, it must be the case that $\sV_{h} \nsubseteq \Union_{h' \neq
    h \in \sG'} \sV_{h'}$ and hence there exists a unique view for $h$.

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the subgraph from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sG'} \sV_{h'}$
\end{proof}

>>>>

In \sectionref{directed}, we stated that bottlenecks guaranteed
exclusive views provided conditional independence assumptions \assumptionref{ci}. 
We provide the proof below.
  We generalize the lemma so that it applies
  on any connected subgraph, not just cliques.
  Also, the statement is made in terms of subgraphs,
  not the entire graph.

\begin{lemma*}[Bottlenecks guarantee exclusive views]
  %\label{lem:bottleneck-views}  
  Let $\sG$ be a graphical model satisfying conditional independence
  assumptions (\assumptionref{ci}) and be uniformly bottlenecked, then
  $\sG$ has the exclusive views property.
%Let $\sG' = \{ h_{i_1}, \ldots, h_{i_m} \} \subseteq \sG$ be
%  a connected subgraph of hidden variables. If every hidden variable
%  $h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive view property.
\end{lemma*}

\begin{proof}
Let $h_0$ be a hidden variable in the clique $\sC$ and $x_1, x_2$ be any
  two conditionally independent observed variables. 
One of $x_1$ or $x_2$ is an exclusive view for $h_0$ iff it is
  independent of all other hidden variables in the clique, given $h_0$.
Intuitively, if this weren't the case, then we would be able to find
  an active path between $x_1$ and $x_2$ that does not go through $h_0$,
  violating the conditional independence condition.

% TODO: One line here about how assumptions save us.
% The conditional independence assumptions (\assumptionref{ci}) exclude
%   edge cases wherein the active trail still preserves 
% for them to be both depend on any other
% variable $h'$ in the clique.

Suppose neither $x_1$ nor $x_2$ were exclusive views for $h_0$, 
  then there exists some $h_1, h_2 \in \sC$ (not necessarily distinct)  that $x_1, x_2$ are
  respectively dependent on:
  $x_1 \not\perp h_1 \given h_0$ and $x_2 \not\perp h_2 \given h_0$. 
Both $h_1$ and $h_2$ are part of the same clique $\sC$, thus by
  \assumptionref{ci}, we have that $h_1 \not\perp h_2 \given h_0$.

Since $x_1 \not\perp h_1 \given h$, there exists an active trail from
$x_1 \to h_1$. Furthermore, because $h_1$ and $h_2$ are in the same
clique, the active trail extends to $h_2$, i.e. $x_1 \not\perp h_2$.
However, since $x_1 \to h_1 \to h_2$ is an active trail in the unconditioned graph, i.e. $x_1 \not\perp h_1$; by \assumptionref{ci}, it remains active trail when conditioned on $h$.

Consequently, $x_1 \not\perp h_2 \given h_0$, which
contradicts the assumption that as $x_1 \perp x_2 \given h_0$,
$x_1 \not\perp h_i \given h_0$ and $x_2 \not\perp h_i \given h_0$ for all $h_i \in \sC$.

\end{proof}
